---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bcps

<!-- badges: start -->
<!-- badges: end -->

The goal of bcps is to make data from Baltimore City Public Schools more consistent and accessible to R users. This package may pair well with the [mapbaltimore package](https://github.com/elipousson/mapbaltimore) that offers a broader range of Baltimore-specific datasets and functions for working with that data.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("elipousson/bcps")
```
## Example

Currently, this package includes datasets that include school and grade-level enrollment and demographic data, the published results from a parent survey, and the published results from a combined student and educator survey completed in 2019. This data can be used to answer questions, such as, what are the elementary schools with the greatest total student enrollment?

```{r enrollment_demographics}
library(bcps)
library(dplyr)

top_5_es <- enrollment_demographics_SY1920 %>% 
  filter(grade_range == "All Grades",
         grade_band == "E") %>% 
  select(school_number, school_name, total_enrollment) %>% 
  top_n(5, total_enrollment) %>% 
  arrange(desc(total_enrollment))

top_5_es %>% 
  knitr::kable(caption = "Five largest BCPS elementary schools by total enrollment")
```

Both the enrollment/demographic data and the parent survey are available in both a wide and long format.

The package also includes spatial data for elementary school attendance zones and program locations for the 2020-2021 school year.

```{r bcps_es_zones}
library(ggplot2)

bcps_es_zones_SY2021 %>% 
  ggplot() +
  geom_sf(aes(fill = zone_name)) +
  scale_fill_viridis_d() +
  guides(fill = "none") +
  labs(title = "BCPS Elementary School Attendance Zones") +
  theme_minimal()
```

These two sources can be used in combinations by joining the `program_number` in the spatial data with the equivalent `school_number` used in the survey and demographic data.

```{r top_5_es_map}
top_5_es_map <- bcps_programs_SY2021 %>% 
  left_join(top_5_es, by = c("program_number" = "school_number")) %>% 
  filter(!is.na(total_enrollment)) %>% 
  ggplot() +
  geom_sf(data = bcps_es_zones_SY2021, fill = NA, color = "darkblue") +
  geom_sf(aes(color = school_name)) +
  geom_sf_label(aes(label = program_name_short, fill = school_name), color = "white") +
  scale_fill_viridis_d(end = 0.7) +
  guides(fill = "none", color = "none") +
  labs(title = "Five largest BCPS elementary schools by total enrollment") +
  theme_minimal()

top_5_es_map
```
